Assessing linkages between watershed nutrient loading and estuary water quality in Lavaca Bay, Texas

### 2023-02-09 Schramm-Manuscript-2023/Schramm-Manuscript-2023.tex
results

Because streamflow and nutrient loads are tightly correlated, freshwater
inflow masks any potential signals from nutrient loads alone. Following
the methodology implemented by
\citet{murphyNutrientImprovementsChesapeake2022}, both streamflow and
nutrient loads were prepossessed to account for season and flow. Instead
of using raw freshwater inflow and nutrient loading values, these values
were replaced by seasonally adjusted inflow and flow-adjusted nutrient
loads by fitting a GAM relating season (day of year) to log transformed
daily freshwater inflow values:

@@

Lavaca NO\textsubscript{3} r2 = 0.85 and 0.90 deviance explained TP r2 =
0.27 and 0.33 Texana NO\textsubscript{3} r2 = 0.75 and 0.81 TP = 0.32
and 0.38

@@

@@

\caption[Aggregated estimated annual and flow-normalized annual NO~3~ and TP loads for USGS-08164000 and USGS-08164525]{Aggregated estimated annual and flow-normalized annual NO~3~ and TP loads for USGS-08164000 and USGS-08164525.}\label{fig:fig3}
\end{figure}

@@

\begin{verbatim}
## Usually it is recommended to use column_spec before collapse_rows, especially in LaTeX, to get a desired result. 
## Usually it is recommended to use column_spec before collapse_rows, especially in LaTeX, to get a desired result. 
## Usually it is recommended to use column_spec before collapse_rows, especially in LaTeX, to get a desired result.
\end{verbatim}

@@

\caption{\label{tab:unnamed-chunk-1}Estuary GAM AIC\textsubscript{c} values and associated model probabilities (in parenthesis). Models with the highest probability for each site and water quality parameter combination are bolded and italicized for emphasis.}
\centering
\begin{tabular}[t]{ll>{}l>{}l>{}l}
\toprule
Parameter & Site & Temporal & Flow & Flow + Load\\
\midrule
 & TCEQ-13383 & -152.1 (0.03) & -156.1 (0.24) & \em{\textbf{-158.2 (0.72)}}\\

@@

{\centering \includegraphics[width=1\linewidth]{Schramm-Manuscript-2023_files/figure-latex/long-smooths-1}

@@

\end{adjustwidth}\caption[Smoothed trend component for water quality paramaters obtained from Temporal estaury GAMs]{Smoothed trend component for water quality paramaters obtained from Temporal estaury GAMs.}\label{fig:long-smooths}
\end{figure}

@@

\begin{tabular}[t]{lllll}
\toprule
Parameter & \makecell[c]{Reported Yield\\(kg$\cdot$km\textsuperscript{-2}$\cdot$year\textsuperscript{-1})} & Approach & Time Period & Reference\\
\midrule
TP & 42.9 (34.4, 54.0) & GAM & 2000-2020 & This work\\
TP & 45.2 & SPARROW & 2012 & \citet{wise_spatially_2019}\\
TP & 42 & SWAT & 1977-2005 & \citet{omaniEstimationSedimentNutrient2014}\\
TP & 20.81-91.58 & SPARROW & 2002 & \citet{rebich_sources_2011}\\
TP & 28.9 & LOADEST & 1972-1993 & \citet{dunnTrendsNutrientInflows1996}\\
\bottomrule
\end{tabular}
\end{table}

>>>>

Because streamflow and nutrient loads are tightly correlated, freshwater
inflow can mask signals from nutrient loads alone. Following the
methodology implemented by
\citet{murphyNutrientImprovementsChesapeake2022}, both streamflow and
nutrient loads were prepossessed to account for season and flow. Instead
of using raw freshwater inflow and nutrient loading values, these values
were replaced by seasonally adjusted inflow and flow-adjusted nutrient
loads by fitting a GAM relating season (day of year) to log transformed
daily freshwater inflow values:

@@

Based on criteria provided by \citet{moriasiHydrologicWaterQuality2015},
GAMs ranged from ``satisfactory'' to ``very good'' based on median NSE,
r\textsuperscript{2} and PBIAS calculated from repeated 5-fold cross
validation on predicted and measured nutrient loads. NO\textsubscript{3}
GAM models had median NSE values of 0.34 and 0.48 at USGS-08164000 and
USGS-08164390 respectively. Median r\textsuperscript{2} values were 0.70
(USGS-08164000) and 0.87 (USGS-08164525) and PBIAS values were 2.00
(USGS-08164000) and 10.90 (USGS-08164525) for NO\textsubscript{3} loads.
Median NSE values were 0.80 (USGS-08164000) and 0.91 (USGS-08164525) for
TP loads. Median r\textsuperscript{2} values for TP loads were 0.93
(USGS-08164000) and 0.99 (USGS-08164525) and median PBIAS values were
-7.20 (USGS-08164000) and -3.30 (USGS-08164525). Density plots of
metrics show similar distribution of values between sites for the same
parameter, with the exception r\textsuperscript{2} values for
NO\textsubscript{3} loads where USGS-08164000 showed a much larger
variance in values compared to USGS-08164525 (Figure \ref{fig:fig2}).In
addition to higher average NSE and r\textsuperscript{2} values, GAMs had
smaller variance in metric values for TP compared to
NO\textsubscript{3}.

@@

Annual NO\textsubscript{3} and TP loads show considerable variation,
generally following patterns in discharge (Figure \ref{fig:fig3}, Figure
\ref{fig:fig4}). Flow-normalized TP loads at both sites and
flow-normalized loads at USGS-08164000 indicate watershed based loads
have not changed much over time when accounting for changes in
streamflow (Figure \ref{fig:fig3}). Flow-normalized loads at
USGS-08164000 show small variation over time with some decreases in
NO\textsubscript{3} loads since 2013. Aggregated across both sites, the
mean annual NO\textsubscript{3} load 2005 through 2020 was 205,405 kg
(126,867 kg - 341,569 kg, 90\% CI). Annual NO\textsubscript{3} loads
ranged from 12,574 kg in 2011 to 794,510 kg in 2007.

Total annual TP loads ranged from 7,839 kg in 2011 to 595,075 kg in
2007. Mean annual TP loading from 2005 through 2020 was 182,673 kg
(152,227 kg - 219,310 kg, 90\% CI). On average, USGS-08164525 accounted
for 68\% of NO\textsubscript{3} loads and 59\% of TP loads from 2005
through 2020. However, during periods of extreme drought the Lavaca
River (USGS-08164000) became the primary source of nutrient loading in
the watershed with the Navidad River only accounting for 15\% and 25\%
of NO\textsubscript{3} and TP loads in 2011 (Figure \ref{fig:fig4}).

@@

\caption[Aggregated estimated annual and flow-normalized annual NO  extsubscript{3} and TP loads for USGS-08164000 and USGS-08164525]{Aggregated estimated annual and flow-normalized annual NO extsubscript{3} and TP loads for USGS-08164000 and USGS-08164525.}\label{fig:fig3}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{Schramm-Manuscript-2023_files/figure-latex/fig5-1}

\end{adjustwidth}\caption[Smoothed temporal trend component for water quality paramaters obtained from temporal estuary GAMs]{Smoothed temporal trend component for water quality paramaters obtained from temporal estuary GAMs.}\label{fig:fig5}
\end{figure}

GAMs did not identify significant changes in TP or DO concentrations at
any of the Lavaca Bay sites from 2005 through 2020 (Figure
\ref{fig:fig5}). The upper-bay site, TCEQ-13563, had a linear increase
in NO\textsubscript{x} concentration and and decrease in
chlorophyll-\emph{a} from 2005 through 2014. The mid-bay site,
TCEQ-13383, showed a periodic pattern in NO\textsubscript{x}
concentration that appeared similar to precipitation/inflow patterns, as
well as a post 2011 increase in TKN concentrations. No significant
long-term trends in concentrations were identified by GAMs for the
lower-bay TCEQ-13384 site.

Freshwater inflow provided additional explanation for changes in TP and
NO\textsubscript{\emph{x}} concentration at all of the Lavaca Bay sites
according to AIC\textsubscript{c} and model probability values (Table
\ref{tab:tab4}). TCEQ-13563, the site closest to the river outlet, was
the only site that had improvements in the explanations of DO and TKN
concentration with the inclusion of inflow. Both TCEQ-13563 and
TCEQ-13383, the mid-bay site, saw improvements in explanations for
variations in chlorophyll-\emph{a} with the inclusion of freshwater
inflow. The addition of nutrient loads (both TP and NO\textsubscript{3})
terms did not provide additional explanation for changes in
chlorophyll-\emph{a} or DO concentrations. Inclusion of TP loads
provided additional explanation of TP concentrations at the upper- and
mid-bay sites, TCEQ-13563 and TCEQ-13383. Inclusion of
NO\textsubscript{3} loads only provided marginal improvements in the
explanation of NO\textsubscript{\emph{X}} concentration at the lower-bay
TCEQ-13384 site.

@@

\caption{\label{tab:tab4}Estuary GAM AIC\textsubscript{c} values and associated model probabilities (in parenthesis). Models with the highest probability for each site and water quality parameter combination are bolded and italicized for emphasis.}
\centering
\begin{tabular}[t]{ll>{}l>{}l>{}l}
\toprule
Parameter & Site & Temporal & Flow & Flow + Load\\
\midrule
 & TCEQ-13383 & -152.1 (0.03) & -156.1 (0.24) & \em{\textbf{-158.2 (0.72)}}\\

@@

{\centering \includegraphics[width=1\linewidth]{Schramm-Manuscript-2023_files/figure-latex/fig6-1}

@@

\end{adjustwidth}\caption[Estimated effects of mean daily inflow residuals on mean TP, NO\textsubscript{\emph{x}}, chlorophyll-\emph{a}, TKN, and DO concentrations in Lavaca Bay obtained from flow estuary GAMs]{Estimated effects of mean daily inflow residuals on mean TP, NO\textsubscript{\emph{x}}, chlorophyll-\emph{a}, TKN, and DO concentrations in Lavaca Bay obtained from flow estuary GAMs.}\label{fig:fig6}
\end{figure}

GAMs show that increases in freshwater inflow result in nearly linear
increases in TP and NO\textsubscript{\emph{x}} at all three sites
(Figure \ref{fig:6}). At the upper-bay TCEQ-13563 site, GAMs showed
increases in freshwater inflow initially increased chlorophyll-\emph{a}
and DO concentration but concentrations leveled and potentially
decreased at higher flows. The mid-bay TCEQ-13383 site showed a nearly
linear increased in chlorophyll-\emph{a} concentration with increases in
freshwater inflow. Freshwater flow did not have significant effects on
chlorophyll-\emph{a}, TKN, or DO at the lower-bay TCEQ-13384 site.

Increases in TP loads resulted in nearly linear increases of TP
concentration at the upper- and mid-bay sites, TCEQ-13563 and TCEQ-13383
respectively (Figure \ref{fig:fig7}). The relative effect size appeared
to much smaller than the effect of freshwater inflow alone. Increases in
NO\textsubscript{3} loads only showed an effect at the lower-bay
TCEQ-13384 site. The effect was quite small compared to streamflow and
provided only small improvements to the model (Table \ref{tab:tab4}). As
noted above, nutrient loadings did not provide any explanation in
changes in the remaining assessed water quality parameters.

{\centering \includegraphics[width=1\linewidth]{Schramm-Manuscript-2023_files/figure-latex/fig7-1}

\caption[Estimated effects of nutrient load residuals on TP and NO\textsubscript{\emph{x}} concentrations in Lavaca Bay obtained from flow+load estuary GAMs]{Estimated effects of nutrient load residuals on TP and NO\textsubscript{\emph{x}} concentrations in Lavaca Bay obtained from flow+load estuary GAMs.}\label{fig:fig7}
\end{figure}

@@

\begin{threeparttable}
\caption{\label{tab:tpcomp}Mean estimates of annual TP yield in the Lavaca River watershed in published studies.}
\centering
\begin{tabular}[t]{lllll}
\toprule
Parameter & \makecell[c]{Reported Yield\\(kg$\cdot$km\textsuperscript{-2}$\cdot$year\textsuperscript{-1})} & Approach & Time Period & Reference\\
\midrule
TP & 35.2 (28.8, 43.3)\textsuperscript{*} & GAM & 2005-2020 & This work\\
TP & 45.2 & SPARROW & 2000-2014 & \citet{wise_spatially_2019}\\
TP & 42 & SWAT & 1977-2005 & \citet{omaniEstimationSedimentNutrient2014}\\
TP & 20.81-91.58\textsuperscript{\dag} & SPARROW & 1980-2002 & \citet{rebich_sources_2011}\\
TP & 28.9 & LOADEST & 1972-1993 & \citet{dunnTrendsNutrientInflows1996}\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item [*] Values represent the mean of annual point estimates, lower and upper 95\% credible intervals.
\item [\dag] A single point estimate was not reported, these value represent the range depicted on the choropleth map provided in the report.
\end{tablenotes}
\end{threeparttable}
\end{table}
